Experience Report: Haskell in Computational Biology

### 2012-06-29 mrfy_experience_report.tex
more on search

Once the successors have been computed and tagged,
\texttt{span} finds the first useful successor.
In~case there \emph{is} no successor, \texttt{everyPt} also returns
all the useless successors.
\remark{Story about the killer strict Rand monad? Where?}
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.
Because \texttt{everyPt} is monadic, the points accumulated so far are
appended to its result using the \verb+<$>+~operator.
%
The most informative part of \texttt{everyPt} is last expression of
the \texttt{do} block,
which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with another useful point.

When the computation returned by \texttt{everyPt} is evaluated against
a random-number generator, it returns an infinite list containing a
mix of useful and useless points.
\remark{This part is gross and still needs work!}
This mix can be passed to a test function of type \texttt{SearchStop~pt}:\\
\vbox{
\smallverbatiminput{strat}
\smallverbatiminputx{stop}
\smallverbatiminputx{search}
}
\noindent

We have two kinds of non-modularity.
\remark{Still gross}
First, termination has to
look at \emph{all} states, even though we would prefer that it look
only at useful states.
This is because otherwise we risk non-termination:
we are not guaranteed a useful move after any finite number of steps.
Second, age of the search shows up in multiple places.
In particular, we cannot separate the age of the search at the
current state from the utility decision.

The \texttt{SearchStrategy} interface provides ample scope for
experiments.\remark{\P\ needs rewriting}
Random-mutation hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.

>>>>

Once the successors have been computed and tagged,
\texttt{span} finds the first useful successor.
In~case there \emph{is} no successor, \texttt{everyPt} also returns
all the useless successors.
\remark{Story about the killer strict Rand monad? Where?}
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.
(Because \texttt{everyPt} is monadic, the points accumulated so far are
appended to its result using the \verb+<$>+~operator.)
%
The most informative part of \texttt{everyPt} is last expression of
the \texttt{do} block,
which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with the results of \texttt{everyPt}.

The rest of the search uses Hughes's
classic composition of generator  and test function.
Because our code is monadic, we~use the monadic composition
operator~\texttt{=<<}, which is the bind operator with its arguments
swapped: 
\smallverbatiminput{search}
The \texttt{test} function has type \mono{SearchStop~pt}:\\
\smallverbatiminputx{stop}
Type \mono{History~pt} retains only the useful points.
(Internally, \mrfy\ needs only the \emph{final} useful point,
but because we want to study how different search algorithms behave,
we~keep all the useful points.)

The definition of \texttt{SearchStop} reveals the forms of
non-modularity which are inherent in \mrfy's search algorithm.
First, we need \texttt{Utility} because if we omit the useless states,
search might not terminate.
Second, we need \texttt{CCosted} because some of our test functions
decide to terminate based either on the cumulative cost of the most
recent point or on the difference between costs of successive points.

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.
